Supporting online material for\\ \emph{Effectiveness of UNAIDS targets and HIV vaccination across 127 countries}

### 2016-08-16 Documents/supplementary_text.tex
A little work.

Projections were generated for each country through model simulations
using 1000 random samples from published parameter distributions
(table S1, Supplementary text) and the country-specific estimated
distribution of the transmission rate. We evaluated the global and
country-level impacts of meeting the UNAIDS targets and vaccine
deployment on four outcome measures: HIV incidence, PLHIV, PLAIDS, and
AIDS-related deaths between 2015 and 2035. Our base-case scenario
assumed maintenance of the status quo: that 2015 levels of the
proportion of PLHIV diagnosed (pD), proportion of those diagnosed on
ART (pT), and proportion of those on treatment who have achieved viral
suppression (pV) continue unchanged until 2035. The intervention
scenarios were (1.) the attainment of the 90–90–90 target by linear
increases in pD, pT, and pV until 2020 followed by maintenance of
those levels until 2035, and (2.) linear increase in the diagnosis and
treatment levels to the 90–90–90 target by 2020, a linear increase
from 90–90–90 to the 95–95–95 target by 2030, and maintenance of the
95–95–95 levels to 2035. (For 90–90–90, if any initial level was above
90\%, it was kept constant going forward so that the intervention did
not worsen coverage.  Likewise, for 95–95–95, any initial level above
95\% was kept constant.)  For vaccination, we assumed deployment of a
vaccine with 50\% efficacy in 2020 with coverage increasing linearly
from 0\% to 50\% at 2 years post-rollout, continuing on the same
linear trend to 70\% at 2.8 years post-rollout, and maintaining
constant 70\% coverage thereafter. We also performed a partial rank
correlation coefficient (PRCC) sensitivity analysis for each model
parameter and a scenario analysis to account for uncertainty in the
profile of the prospective vaccine. See Supplementary text for more
details.

@@

The PLHIV population was comprised of people in the acutely infected,
undiagnosed, diagnosed, treated, viral suppression and AIDS
compartments. People susceptible to HIV (S) progress to the
acute-infection phase (A) according to the force of infection, which
depends on the HIV transmission rate, estimated from the
country-specific incidence and prevalence data (next section),
adjusted for the current number of acutely infected, untreated or
ineffectively treated, and viral suppression PLHIV.  Acute infection
is characterized by high viral loads and lasts 88 days on average
\cite{Hollingsworth2008-iy}. We assumed that individuals remain
undiagnosed during their acute infection phase and during at least
some of the subsequent period of clinical latency (i.e., the
undiagnosed compartment). Individuals in the undiagnosed class will be
diagnosed at a rate determined by the country's initial diagnosis
level and potential progress towards the UNAIDS targets. Diagnosed
people transition to the treated compartment upon starting ART at a
rate of $\phi$ , and to the virally suppressed class (V) after
achieving viral suppression through treatment retention at a rate of
$\gamma$. The disengagement from treatment moves individuals (T, V) at
a rate of $\omega$ back to the diagnosed class (D). Rates of movement
between the undiagnosed (U), diagnosed (D), treated (T) and virally
suppressed classes (V) are determined by the proportions of PLHIV
diagnosed ($p_{D}$), proportion of diagnosed individuals on ART
($p_{T}$), and the proportion of individuals on ART who have achieved
viral suppression ($p_{V}$). PLHIV in U, D, T, and V classes develop
AIDS (W) after an average duration dependent on viral suppression
status. Individuals who have achieved viral suppression (V) take
longer to develop AIDS ($1/\theta$), than those who are in the other
PLHIV classes ($1/\sigma$). Individuals who are susceptible to HIV can
be vaccinated at a of rate ($\zeta$), and depending on vaccine
efficacy, have a reduced chance of HIV infection. The model equations
are

\begin{equation}
  \label{model_eqns}
  \begin{split}
    \frac{\md S}{\md t} &= B N - \lambda S - \zeta S- \mu S,
    \\
     \frac{\md R}{\md t} & = \zeta S - \rho \lambda R - \mu R,
    \\
    \frac{\md A}{\md t} &= \lambda S +\rho \lambda R - \delta A - \mu A,
    \\
    \frac{\md U}{\md t} &= \delta A - \alpha U - \mu U - \sigma U,
    \\
    \frac{\md D}{\md t} &=  \alpha U + \psi T + \psi V
    - \phi D - \mu D - \sigma D,
    \\
    \frac{\md T}{\md t} &= \phi D - \psi T - \gamma T - \mu T
    - \sigma T,
    \\
    \frac{\md V}{\md t} &= \gamma T - \psi V - \mu V - \theta V,
    \\
    \frac{\md W}{\md t} &= \sigma U + \sigma D + \sigma T + \theta V -
    \nu W,
  \end{split}
\end{equation}
with force of infection
\begin{equation}
  \label{force_of_infection}
  \begin{split}
    \lambda &= \frac{\beta_A A + \beta_U (U + D + T) + \beta_T V}{N},
    \\
    N &= S + R +  A + U + D + T + V,
    \\
    \beta_x &= c \left[1 - (1 - \tau_x)^n\right]
    \text{ for $x \in \{A, U, T\}$},
    \\
    \tau_T &= (1 - \epsilon) \tau_U.
  \end{split}
\end{equation}

The rate of developing AIDS  among those who have acheived
viral suppression ($\theta$) is
\begin{equation}
 \frac{1}{1/\mu-\omega-1/\nu}.
\end{equation}

The proportion diagnosed is
\begin{equation}
  p_D = \frac{D + T + V + W}{A + U + D + T + V + W},
\end{equation}
the proportion treated is
\begin{equation}
  p_T = \frac{T + V + W}{D + T + V + W},
\end{equation}
and the proportion with viral suppression is
\begin{equation}
  p_V = \frac{V}{T + V}.
\end{equation}

@@

Similarly, the proportion of people vaccinated is
\begin{equation}
P_{V} = \frac{R}{S+R},
\end{equation}

We can implement the vaccination compartment in the same way as the other controls. When we begin
vaccination in 2020/2025, the initial vaccination level is 0\%. This
coverage will increase linearly to $50\%$ within first 2 or 5
years. The coverage will continue to increase linearly till it reaches
the desired coverage ($50\%,70\%,90\%$) and then stay constant for the
remainder of the projections.

95--95--95 goal is implemented subsequently after meeting 90--90--90
by 2020 by increasing $p_D$, $p_T$, and $p_V$ to 95--95--95 levels by
2030 linearly. \comment{@Jan, We use the Heaveside function here too?}

\subsection{Rates that are on when $p_x < p^*_x$}

@@

\section{Model fitting}
\label{model_fitting}
We first estimated country-specific rates of HIV transmission for each
of the available historical data point for incidence and
prevalence. The calculation involved a simplified force of infection
that initially assumed no differences in transmission risk among
acute, unsuppressed, suppressed, and AIDS classes due to the general
lack of historical data on diagnosis and treatment. We calculated a
time-independent aggregate transmission rate by taking an exponential
weighted mean of the historical (i.e., time-dependent) transmission
rates for each given country (\autoref{transmission_rate}). The
aggregate transmission rate was then combined with estimated HIV
transmission risk by stage of HIV infection
\cite{Hollingsworth2008-iy} and treatment status \cite{Wawer2005-us}
to calculate separate transmission rates for unsuppressed, acute, and
viral-suppression people. These rates were used to simulate model
projections through 2035.

@@

\begin{figure}
  \centering
  \includegraphics{../Codes/plots/transmission_rate.pdf}
  \caption{Country-specific transmission rates were calculated based
    on the relative contributions of individuals with acute infection,
    unsuppressed viral load or suppressed viral load to overall
    transmission (\autoref{model_fitting}). The estimates were
    calibrated to data on country-level incidence and/or prevalence to
    account for variation in historical incidence while accurately
    reflecting recent HIV rates.}
  \label{transmission_rate}
\end{figure}

@@

We evaluated the global and country-specific impacts of meeting the
UNAIDS targets and vaccine deployment on four outcome measures: HIV
incidence, PLHIV, number of people with AIDS, and AIDS-related deaths
between 2015 and 2035. Calibrating our model to historical data on
prevalence and incidence, we projected the anticipated trajectories of
the outcomes using country-specific transmission rates across the
intervention scenarios outlined below.  Our base-case scenario assumed
maintenance of the status quo: that 2015 levels of the proportion of
PLHIV diagnosed ($p_{D}$), proportion of those diagnosed on ART
($p_{T}$), and proportion of those on treatment who have achieved
viral suppression ($p_{V}$) continue unchanged until 2035.  The
intervention scenarios were (1.) the attainment of the 90--90--90
target by linear increases in $p_{D}$, $p_{T}$, and $p_{V}$ until 2020
followed by maintenance of those levels until 2035, and (2.) linear
increase in the diagnosis and treatment levels to the 90--90--90
targets by 2020, a linear increase from 90--90--90 to the 95--95--95
target by 2030, and maintenance of the 95--95--95 levels to 2035. (For
90--90--90, if any initial level was above 90\%, it was kept constant
going forward so that the intervention did not worsen coverage.
Likewise, for 95--95--95, any initial level above 95\% was kept
constant.)  For vaccination, we assumed deployment of a vaccine with
50\% efficacy in 2020 with coverage increasing linearly from 0\% to
50\% at 2 years post-rollout and maintaining constant coverage after
reaching 70\%. The added benefit of vaccination was considered in the
context of status quo diagnosis and treatment, achieving the UNAIDS
90–90–90 goal by 2020 followed by maintenance, and achieving the
95–95–95 goal by 2030 followed by maintenance.  Projections were
generated through model simulations based on published parameter
distributions (\autoref{model_param}) and country-specific estimated
transmission rate distributions. The parameters were sampled using
latin hypercube sampling \cite{}
and the variance reduction technique of running different
interventions with random parameter values \cite{}.

We calculated partial rank correlation coefficients (PRCC) to measure
the independent effect of each parameter on model projections of HIV
incidence (\autoref{PRCCs}). Furthermore, due to uncertainty
surrounding the prospective vaccine, we obtained model projections
using six additional vaccine scenarios by considering alternative
estimates of efficacy (30\%, 70\%), coverage (50\%, 90\%), time to
50\% coverage (2 years, 5 years), and first year of availability
(2025).

@@

\subsection{Effectiveness of HIV interventions globally, regionally
  and for each of 127 countries}

Model outcomes were projected until 2035 for each of the 127 countries
with sufficient data on status quo rates of diagnosis and treatment
and on historical incidence or prevalence. The median estimates with
interquartile ranges are presented between 2015 and 2035 for number of
PLHIV, annual incidence (per million population), number of people
with AIDS, and number of HIV-related deaths under the status quo and
five intervention scenarios: vaccination only, attainment of the
UNAIDS 90--90--90 target, 90--90--90 target and vaccination, attainment of
the UNAIDS 95--95--95 target, and 95--95--95 target and vaccination.  The
country-specific projections as well as aggregated projections at the
regional and global levels follow:

>>>>

Projections were generated for each country through model simulations
using 1000 random samples from published parameter distributions
(table S1, Supplementary text) and the country-specific estimated
distribution of the transmission rate. We evaluated the global and
country-level impacts of meeting the UNAIDS targets and vaccine
deployment on four outcome measures: HIV incidence, PLHIV, PLAIDS, and
AIDS-related deaths between 2015 and 2035. Our base-case scenario
assumed maintenance of the status quo: that 2015 levels of the
proportion of PLHIV diagnosed (pD), proportion of those diagnosed on
ART (pT), and proportion of those on treatment who have achieved viral
suppression (pV) continue unchanged until 2035. The intervention
scenarios were (1.) the attainment of the 90–90–90 target by linear
increases in pD, pT, and pV until 2020 followed by maintenance of
those levels until 2035, and (2.) linear increase in the diagnosis and
treatment levels to the 90–90–90 target by 2020, a linear increase
from 90–90–90 to the 95–95–95 target by 2030, and maintenance of the
95–95–95 levels to 2035. (For 90–90–90, if any initial level was above
90\%, it was kept constant going forward so that the intervention did
not worsen coverage.  Likewise, for 95–95–95, any initial level above
95\% was kept constant.)  For vaccination, we assumed deployment of a
vaccine with 50\% efficacy in 2020 with coverage increasing linearly
by 25\% per year, continuing up to 70\% at 2.8 years post-rollout, and
maintaining constant 70\% coverage thereafter.  We also performed a
partial rank correlation coefficient (PRCC) sensitivity analysis for
each model parameter and a scenario analysis to account for
uncertainty in the profile of the prospective vaccine.

@@

The PLHIV population was comprised of people in the acutely infected,
undiagnosed, diagnosed, treated, viral suppression and AIDS
compartments. People susceptible to HIV ($S$) progress to the
acute-infection phase ($A$) according to the force of infection, which
depends on the HIV transmission rate, estimated from the
country-specific incidence and prevalence data
(\autoref{model_fitting}), adjusted for the current number of acutely
infected, untreated or ineffectively treated, and viral suppression
PLHIV.  Acute infection is characterized by high viral loads and lasts
88 [XX, YY] days \cite{Hollingsworth2008-iy}. We assumed that people
remain undiagnosed during their acute infection phase and during at
least some of the subsequent period of clinical latency (i.e., the
undiagnosed compartment). After acute infection, people move into the
chronic undiagnosed class ($U$).  People in the undiagnosed class will
be diagnosed at a rate determined by the country's initial diagnosis
level and potential progress towards the UNAIDS targets.  Diagnosed
people (compartment $D$) transition to the treated compartment ($T$)
upon starting ART at a rate of $\phi$ , and then to the
viral-suppression class ($V$) after achieving viral suppression
through treatment retention at a rate of $\gamma$. The disengagement
from treatment moves people from treated ($T$) and viral suppression
($V$) at a rate of $\omega$ back to the diagnosed class ($D$). Rates
of movement between the undiagnosed ($U$), diagnosed ($D$), treated
($T$) and viral-suppression classes ($V$) are determined by the
proportions of PLHIV diagnosed ($p_{D}$), proportion of diagnosed
people on ART ($p_{T}$), and the proportion of people on ART who have
achieved viral suppression ($p_{V}$). PLHIV in $U$, $D$, $T$, and $V$
classes develop AIDS (compartment $W$) after an average duration
dependent on viral suppression status. People who have achieved viral
suppression ($V$) take longer to develop AIDS ($1/\theta$), than those
who are in the other PLHIV classes ($1/\sigma$). People who are
susceptible to HIV are vaccinated at rate $\zeta$, transitioning into
the vaccinated compartment ($R$), where the vaccine reduces their
force of infection by the efficacy factor $\rho$.  The model equations
are
\begin{equation}
  \label{model_eqns}
  \begin{split}
    \frac{\md S}{\md t} &= B N - \lambda S - \zeta S- \mu S,
    \\
    \frac{\md R}{\md t} & = \zeta S - (1 - \rho) \lambda R - \mu R,
    \\
    \frac{\md A}{\md t} &= \lambda S + (1 - \rho) \lambda R - \delta A - \mu A,
    \\
    \frac{\md U}{\md t} &= \delta A - \alpha U - \mu U - \sigma U,
    \\
    \frac{\md D}{\md t} &=  \alpha U + \psi T + \psi V
    - \phi D - \mu D - \sigma D,
    \\
    \frac{\md T}{\md t} &= \phi D - \psi T - \gamma T - \mu T
    - \sigma T,
    \\
    \frac{\md V}{\md t} &= \gamma T - \psi V - \mu V - \theta V,
    \\
    \frac{\md W}{\md t} &= \sigma U + \sigma D + \sigma T + \theta V -
    \nu W,
  \end{split}
\end{equation}
with force of infection
\begin{equation}
  \label{force_of_infection}
  \begin{split}
    \lambda &= \frac{\beta_A A + \beta_U (U + D + T) + \beta_T V}{N},
    \\
    N &= S + R +  A + U + D + T + V,
    \\
    \beta_x &= c \left[1 - (1 - \tau_x)^n\right]
    \text{ for $x \in \{A, U, T\}$},
    \\
    \tau_T &= (1 - \epsilon) \tau_U.
  \end{split}
\end{equation}
The rate of developing AIDS among those who have acheived viral
suppression is
\begin{equation}
 \theta = \frac{1}{1/\mu-\omega-1/\nu}.
\end{equation}

The proportion diagnosed is
\begin{equation}
  p_D = \frac{D + T + V + W}{A + U + D + T + V + W},
\end{equation}
the proportion treated is
\begin{equation}
  p_T = \frac{T + V + W}{D + T + V + W},
\end{equation}
the proportion with viral suppression is
\begin{equation}
  p_V = \frac{V}{T + V},
\end{equation}
and the proportion of people vaccinated is
\begin{equation}
  P_{V} = \frac{R}{S+R}.
\end{equation}

@@

We can implement the vaccination compartment in the same way as the
other controls. When we begin vaccination in 2020 (also 2025 in
scenario analysis), the initial vaccination level is 0\%. The coverage
will increase linearly by 25\% per year (also 10\% per year in
scenario analysis). The coverage will continue to increase linearly
until it reaches 70\% coverage (50\% \& 90\% in scenario analysis) and
then stay constant for the remainder of the time period.

95--95--95 goal is implemented subsequently after meeting 90--90--90
by 2020 by increasing $p_D$, $p_T$, and $p_V$ to 95--95--95 levels by
2030 linearly. \comment{@Jan, We use the Heaveside function here too?
JM: Yes.  I'll write it at some point.}

@@

\section{Model fitting}
\label{model_fitting}

We first estimated country-specific rates of HIV transmission for each
of the available historical data point for incidence and
prevalence. The calculation involved a simplified force of infection
that initially assumed no differences in transmission risk among
acute, unsuppressed, suppressed, and AIDS classes due to the general
lack of historical data on diagnosis and treatment. We calculated a
time-independent aggregate transmission rate by taking an exponential
weighted mean of the historical (i.e., time-dependent) transmission
rates for each given country (\autoref{transmission_rate}). The
aggregate transmission rate was then combined with estimated HIV
transmission risk by stage of HIV infection
\cite{Hollingsworth2008-iy} and treatment status \cite{Wawer2005-us}
to calculate separate transmission rates for unsuppressed, acute, and
viral-suppression people. These rates were used to simulate model
projections through 2035.

@@

\begin{figure}
  \centering
  \includegraphics{../Codes/plots/transmission_rate.pdf}
  \caption{Country-specific transmission rates were calculated based
    on the relative contributions of people with acute infection,
    unsuppressed viral load or suppressed viral load to overall
    transmission (\autoref{model_fitting}). The estimates were
    calibrated to data on country-level incidence and/or prevalence to
    account for variation in historical incidence while accurately
    reflecting recent HIV rates.}
  \label{transmission_rate}
\end{figure}

@@

We evaluated the global and country-specific impacts of meeting the
UNAIDS targets and vaccine deployment on four outcome measures: HIV
incidence, PLHIV, number of people with AIDS, and AIDS-related deaths
between 2015 and 2035. Calibrating our model to historical data on
prevalence and incidence, we projected the anticipated trajectories of
the outcomes using country-specific transmission rates across the
intervention scenarios outlined below.  Our base-case scenario assumed
maintenance of the status quo: that 2015 levels of the proportion of
PLHIV diagnosed ($p_{D}$), proportion of those diagnosed on ART
($p_{T}$), and proportion of those on treatment who have achieved
viral suppression ($p_{V}$) continue unchanged until 2035.  The
intervention scenarios were 1.) the attainment of the 90--90--90
target by linear increases in $p_{D}$, $p_{T}$, and $p_{V}$ until 2020
followed by maintenance of those levels until 2035, and 2.) linear
increase in the diagnosis and treatment levels to the 90--90--90
targets by 2020, a linear increase from 90--90--90 to the 95--95--95
target by 2030, and maintenance of the 95--95--95 levels to 2035. (For
90--90--90, if any initial level was above 90\%, it was kept constant
going forward so that the intervention did not worsen coverage.
Likewise, for 95--95--95, any initial level above 95\% was kept
constant.)  For vaccination, we assumed deployment of a vaccine with
50\% efficacy in 2020 with coverage increasing linearly from 0\% to
50\% at 2 years post-rollout and maintaining constant coverage after
reaching 70\%. The added benefit of vaccination was considered in the
context of status quo diagnosis and treatment, achieving the UNAIDS
90–90–90 goal by 2020 followed by maintenance, and achieving the
95–95–95 goal by 2030 followed by maintenance.  Projections were
generated through model simulations based on published parameter
distributions (\autoref{model_param}) and country-specific estimated
transmission rate distributions. The parameters were sampled using
latin hypercube sampling \cite{}
and the variance reduction technique of running different
interventions with random parameter values \cite{}.

Model outcomes were projected until 2035 for each of the 127 countries
with sufficient data on status quo rates of diagnosis and treatment
and on historical incidence or prevalence
(\hyperref[effectiveness_Global]{\autoref*{effectiveness_Global}}--\ref{effectiveness_Zimbabwe}). The
median estimates with interquartile ranges are presented between 2015
and 2035 for number of PLHIV, per-capita incidence, PLAIDS, and number
of HIV-related deaths under the status quo and five intervention
scenarios: vaccination only, attaining the UNAIDS 90--90--90 target,
attaining the 90--90--90 target and vaccination, attaining the UNAIDS
95--95--95 target, and attaining the 95--95--95 target and
vaccination.  The results from the country were aggregated to the
regional and global levels by adding the numbers of people in each
compartment over time and then PLHIV, per-capita incidence, PLAIDS,
and HIV-related deaths were calculated from the aggregates.

We calculated partial rank correlation coefficients (PRCC) \cite{}
to measure the independent effect of each parameter on model
projections of HIV incidence (\autoref{PRCCs}). Furthermore, due to
uncertainty surrounding the prospective vaccine, we obtained model
projections using six additional vaccine scenarios by considering
alternative estimates of efficacy (30\%, 70\%), coverage (50\%, 90\%),
speed of scale up (10\% coverage per year), and first year of
availability (2025).

@@
